---
title: "Homework 2"
subtitle: "Sample solution"
author: "Boyko Amarov"
date: "4/19/2021"
output:
  bookdown::html_document2: default
---

# Homework 2

- Please write your answers beneath each question and please do not remove the questions from your solution
- Please use the code chunks to write your code
- Make sure that your solution compiles to HTML without errors.
- Copy-paste your solution into the same file from which you downloaded the questions. Please do not create new files in the github repository.


> NOTE that the use of Latex math syntax is not required for your solution.

> NOTE that figure captions and references are not required for your solution. Here they are used for
> better readibility.

```{r}
library(tidyverse)

wines <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/wines2012.csv")
marriage <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/marriage.csv")
```


## Problem 1

The data set `wines` contains scores assigned to 180 wines by several judges.

Variables description:
  
- `score` (numeric): The score assigned to wine i
- `judge.amer` (numeric): Equals 1 if the judge is American and is 0 otherwise.

```{r}
glimpse(wines)
```
1. Fit the following linear model using the `lm` function.

$$
  \text{score}_i = \alpha + \beta \text{judge.amer}_i + e_i, \quad e_i \sim N(0, \sigma^2)
$$


```{r}
## Fit the model here.

fit_wines <- lm(score ~ judge.amer, data = wines)
summary(fit_wines)
```
2. What is the meaning of the coefficients $\alpha$ and $\beta$?

The constant in the model ($\alpha$) is the expected wine score from Non-American judges (`judge.amer = 0`). 
The slope ($\beta$) is the difference between the expected scores from American (`judge.amer = 1`) and Non-American judges (`judge.amer = 0`). In the sample the estimates are $\hat{\alpha} = 13.83$: the sample average score from Non-American judges and $\hat{\beta} = 0.66$: the difference between the sample average scores from American and Non-American judges.


3. Test the null hypothesis against a two-sided alternative that there is no systematic difference between the scores assigned by American and by Non-American judges. Two-sided alternative means that you would reject for both large negative and large positive values of the t-statistic.

- Formulate the hypothesis in terms of the model coefficients.

$$
H_0: \beta = 0\\
H_1: \beta \neq 0
$$
_Hint_: the use of Latex syntax for math is _not_ required for your solution, you can write it in any way as long as it is understandable.

- Write down the value of the t-statistic 
- Decide whether to reject the null hypothesis at both 5% and 10% significance levels (probability of rejection of a true null hypothesis) and explain your decision. Hint: compute the critical values using the `qt` function.


The value of the t-statistic is 1.669 (from the lm summary). The model has two coefficients, therefore the degrees of freedom of the distribution of the t-statistic are $n - p = 180 - 2 = 178$. Using the `qt` function the critical values at both significance levels are:

```{r}
## At 5% significance level
qt(0.0275,df=176)
qt(0.975,df=176)
```

```{r}
## At 10% significance level
qt(0.05,df=176)
qt(0.95,df=176)
```

At the 5% level of significance the null hypothesis cannot be rejected, because the value of the t-statistic (1.669) is
between the two critical values $(-1.93, +1.93)$.


At 10% significance level the null hypothesis should be rejected, because the value of the t-statistic (1.669) is greater than the upper critical value (1.65).



## Problem 2

```{r}
glimpse(marriage)
```

The dataset `marriage` contains information about marriage and divorce rates in 50 US states.

- `Location` (character): Name of the state
- `Marriage` (numeric): Marriage rate in state i
- `MedianAgeMarriage` (numeric): Median age at marriage at first marriage (in years) in state i
- `Divorce` (numeric): Divorce rate in state i

1. Graphical exploration

- Create a scatterplot for the divorce rate (`Divorce`) and the median age at marriage (`MedianAgeMarriage`)
- Create a scatterplot for the divorce rate (`Divorce`) and the marriage rate (`Marriage`)

What patterns of association are visible in the plots?

```{r scatterplot-age-divorce, fig.cap="Scatterplot median age at marriage and divorce rates."}
ggplot(data = marriage, aes(x = MedianAgeMarriage, y = Divorce)) + 
  geom_point() +
  labs(
    y = "Divorce rate",
    x = "Median age at marriage"
  )
```


```{r marriage-divorce, fig.cap="Scatterplot: marriage rates and divorce rates."}
ggplot(data = marriage, aes(x = Marriage, y = Divorce)) + 
  geom_point() +
  labs(
    y = "Divorce rate",
    x = "Marriage rate"
  )
```
From the plots you can see a positive association between marriage rates and divorce rates. States with higher marriage rates also tend to have high divorce rates (see Figure \@ref(fig:scatterplot-age-divorce)).

States that have a low median age at marriage tend to have higher divorce rates (Figure \@ref(fig:marriage-divorce)).


2. Fit the following linear regression model (model 1) and interpret the estimated coefficients

$$
  \text{Divorce}_i = \alpha + \beta \text{MedianAgeMarriage}_i + e_i, e_i \sim N(0, \sigma^2)
$$
  
```{r}
Model1 <- lm(Divorce ~ MedianAgeMarriage, data = marriage)
summary(Model1)
```
The intercept in the model ($\alpha$) corresponds to the expected divorce rate in a state with zero median age at marriage (not realistic).

The slope in the model is the expected difference in divorce rates between states that differ by one year of median age at marriage. The estimated slope ($\hat{\beta} = -0.87$) is negative, indicating that the expected divorce rate is lower in states with higher median age at marriage.


3. Fit the following linear regression model (model 2) and interpret the estimated coefficients

$$
  \text{Divorce}_i = \alpha + \beta \text{Marriage}_i + e_i, e_i \sim N(0, \sigma^2)
$$

```{r}
Model2 <- lm(Divorce ~ Marriage, data = marriage)
summary(Model2)
```
The intercept ($\alpha$) in this model is the expected divorce rate of a state with zero marriage rate (not realistic).
The slope ($\beta$) in this model is the expected difference between the divorce rates of states that differ by one unit of marriage rate. The estimated slope ($\hat{\beta} = 0.17$), indicating that states with higher marriage rate also tend to have a high divorce rate.


4. Fit the following linear regression model (model 3) and interpret the estimated coefficients and interpret the estimated coefficients.

$$
  \text{Divorce}_i = \alpha + \beta_1 \text{Marriage}_i +  \beta_2 \text{MedianAgeMarriage} + e_i, e_i \sim N(0, \sigma^2)
$$
  
```{r}
Model3 <- lm(Divorce ~ Marriage + MedianAgeMarriage, data = marriage)
summary(Model3)
```

5. Explain the change of the coefficient for `Marriage` between model 2 and model 3.

The estimated coefficient of `Marriage` in Model 2 is positive (0.17) and significant (you can reject the null hypothesis that the coefficient is zero: 2.792 >  2.01: the upper critical value at 5% sign. level and 50 - 2 = 48 degrees of freedom).

In Model 3 the coefficient is even negative and not significant at the 5% sign. level (you cannot reject the null hypothesis that the coefficient is zero: -0.706 > -2.01).

```{r}
## Compute the critical values (the difference between the upper and the lower critical values is only the sign)

## Critical values for model 2
qt(0.025, df = 50 - 2)

## Critical values for model 3
qt(0.025, df = 50 - 3)
```

The meaning of the coefficient for `Marriage` is different in the two models. In Model 3 $\beta_1$ is the expected difference in divorce rates for one unit of `Marriage`, but _given_ a fixed age at marriage. The coefficient is close to zero (and not significant). This indicates that knowing the marriage rate of a state helps with predicting the divorce rate if you already know the median age at marriage.

_Extra explanation_: The reason for the change of the coefficient between the two models is that median age at marriage is negatively associated with _both_ the divorce and the marriage rate (Figures \@ref(fig:scatterplot-age-divorce), \@ref(fig:scatterplot-age-marriage)). States with high median age at marriage tend to have both low marriage rates _and_ low divorce rates and the other way around: states with low marriage rates tend to have a high median age at marriage and therefore high divorce rates, see Figure \@ref(fig:scatter-marriage-divorce-age).

```{r scatterplot-age-marriage, fig.cap="Scatterplot: median age at marriage and marriage rates."}
marriage %>%
  ggplot(aes(x = MedianAgeMarriage, y = Marriage)) +
  geom_point() + 
  labs(
    title = "Association between marriage rate and age at marriage",
    x = "Median age at marriage",
    y = "Marriage rate"
  )
```


```{r scatter-marriage-divorce-age, fig.cap="Association between marriage rate, divorce rate and age at marriage. Age at marriage is grouped into two classes (colours) for better visibility."}
## Code for illustration purposes
marriage %>%
  ggplot(aes(x = Marriage, y = Divorce, colour = cut_interval(MedianAgeMarriage, n = 2))) +
  geom_point() + 
  labs(
    title = "Association between marriage rate and age at marriage",
    x = "Marriage rate",
    y = "Median age at marriage",
    colour = "Median age at marriage (two groups)"
  )
```

6. Use model 3 to estimate the expected divorce rate for a state with `MedianAgeMarriage = 25` and `Marriage = 20.3`. Hint: use the `predict` function.

```{r}
predict(Model3, newdata = data.frame(MedianAgeMarriage=25, Marriage=20.3), se.Model3 = TRUE)
```

7. Create new variables `Marriage_centered` and `MedianAgeMarriage_centered` by subtracting the means of the respective variables.

```{r}
marriage <- marriage %>%
  mutate(
    Marriage_centered = Marriage - mean(Marriage),
    MedianAgeMarriage_centered = MedianAgeMarriage - mean(MedianAgeMarriage)
  )
```


8. Fit the following linear regression model (model 4) and interpret the estimated coefficients. Hint: the overline over a variable refers to the mean of this variable.

$$
  \text{Divorce}_i = \alpha + \beta_1 (\text{Marriage}_i - \overline{\text{Marriage}}) +  \beta_2 (\text{MedianAgeMarriage}_i - \overline{\text{MedianAgeMarriage}}) + e_i, e_i \sim N(0, \sigma^2)
$$

```{r}
Model4 <- lm(Divorce ~ Marriage_centered +MedianAgeMarriage_centered, data = marriage)
summary(Model4)
```

9. Explain the change in the estimated intercept in model 4 and in model 5.

The intercept in Model 4 is the expected divorce rate for a state with an average marriage rate and an average median age at marriage (averaged over all 50 states).

10. Test the hypothesis (against a two-sided alternative) that `Marriage` is not linearly related to `Divorce`  given the age at marriage. Formulate the hypothesis in terms of the model coefficients and decide whether to reject it at a 5% significance level (probability of wrongful rejection of a true null hypothesis).


$$
H_0: \beta_1 = 0\\
H_1: \beta_1 \neq 0
$$
The formulation of the statements in terms of the model coefficients derives from the interpretation of the coefficients (see above).

```{r}
## COmpute the critival values (df = 50 states - 3 coefficients)
qt(0.025, df = 50 - 3)
```
The value of the t-statistic (see the lm summary) lies between the two critical values $-0.056 \in (-2.01, 2.01)$. Therefore you cannot reject the null hypothesis at the 5% significance level.
