---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---

## Data import

```{r, setup, message=FALSE}
## Load packages
library(tidyverse)
## Read the data
earnings <- read.csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```


```{r}
glimpse(earnings)
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years

Create a new column holding the height in cm. We use the `mutate` function for this task.

```{r}
##
## earnings <- mutate(earnings, height_cm = 2.54 * height)
## Equivalently using pipes:
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

```{r}
mean(earnings$height_cm)
```
The average height of the persons was 169 cm.

```{r}
summary(earnings$height_cm)
```
```{r}
mean(earnings$height_cm < 162.6)
```
The tallest person was 208 cm tall. The shortest person was 144.8 cm. Median equals 167.6 cm meaning that: _about_ half of the persons
in the survey were shorter than 167.6 cm (median/2nd quartile/0.5 quantile/50% percentile). 1st quartile equals 162.6: _roughly_ (because of rounding in height value) one quarter (25%) were shorter than 162.6 cm. 3th quartile equals 175.9 cm: _about_ 75% (three quarters) of the persons were shorter than 175.9 cm.

## Box-Whiskers plot

```{r}
ggplot(data = earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```
```{r}
ggplot(data = earnings, aes(x = height_cm, y = factor(male))) + geom_boxplot()
```

```{r}
## Extreme case of no variation in the data (all values equal)
summary(c(2, 2, 2, 2, 2, 2, 2))
ggplot(data = data.frame(x = c(2, 2, 2, 2, 2, 2, 2)), aes(x = x)) + geom_boxplot()
```

1. Location: (measured by the median)
2. Spread: (distance between min/max and between the 1st and the 3the quartiles)
3. Outliers: observations that are different from the rest of the observations (persons)

For _categorical_ variables we use frequency tables as a summary.

```{r}
table(earnings$ethnicity)
```

```{r}
ggplot(data = earnings, aes(x = ethnicity)) + geom_bar()
```

There are 180 black persons in the dataset.

## Measurement scales

```{r}
earnings <- earnings %>%
  mutate(
    sex = ifelse(male == 1, "Male", "Female")
  )
```

Categorical scales:

1. **Nominal**: Values of the variable have no numeric meaning. Values are only labels indicating group membership.
2. **Ordinal**: Values of the variable are just labels (as is the case with the nominal scale), but the labels are ordered.
 school grades
 "2" "3" "4" "5" "6"

Numeric scales
<!-- 3. **Interval**: -->
4. **Ratio scale**: Numeric scale, ratios between values make sense.


## Association

Scatter plot

```{r}
NA + 2
```


```{r}
## Compute the mean by simply omitting the missing values
mean(earnings$weight_kg, na.rm = TRUE)

mean(earnings$height_cm)

## NA: missing/undefined value
```


```{r}
ggplot(
  data = earnings, 
  ## aes: aesthetics
  mapping = aes(
    x = weight_kg, 
    y = height_cm
    )
  ) +
  ## Add points
  geom_point(
    ## Add some random noise to all points to avoid overplotting
    position = "jitter",
    ## Transparency
    alpha = 0.2
    ) +
  ## Add horizontal line at the average height (cm)
  geom_hline(yintercept = 169, color = "steelblue2") + 
  ## Add vertical line at the average weight (kg)
  geom_vline(xintercept = 70, color = "firebrick4") +
  ## Add labels
  labs(
    x = "Weight (kg)",
    y = "Height (cm)",
    title = "Scatterplot weight and height"
  )
```
From the points in the scatter-plot we see that heavier persons also tended to be taller persons.

## Sample Covariance


Some notation: 

$$
\text{Avearge values}:\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^n x_i\\
\bar{y} = \frac{1}{n}\sum_{i = 1}^n y_i
$$

Sum of the products over all observation (persons).

$$
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y}) \text{ the sign of this sum depends on the arrangement of the points}
$$
```{r}
## Compute the covariance between height_cm and weight_cm
## Omit missing values for weight: use complete obs instructs it to use only 
## the persons with available height and weight measurements (no missing)

cov(earnings$height_cm, earnings$weight_kg, use = "complete.obs")
```

```{r}
cov(earnings$height, earnings$weight, use = "complete.obs")
```

<!-- $$ -->
<!-- x_i: \text{weight of person i}\\ -->
<!-- y_i: \text{height of person i}\\ -->
<!-- (x_i - \bar{x}) \geq 0 \text{ for persons in upper right quadrant (high-high) }\\ -->
<!-- (y_i - \bar{y}) \geq 0 \text{ for persons in upper right quadrant (high-high) }\\ -->
<!-- (x_i - \bar{x})(y_i - \bar{y}) \geq 0 \text{ for persons in upper right quadrant (high-high) }\\ -->
<!-- (x_i - \bar{x})(y_i - \bar{y}) \geq 0 \text{ for persons in the bottom left quadrant (low-low) }\\ -->
<!-- (x_i - \bar{x})(y_i - \bar{y}) \leq 0 \text{ for persons in the bottom right quadrant (high-low) }\\ -->
<!-- (x_i - \bar{x})(y_i - \bar{y}) \leq 0 \text{ for persons in the upper left quadrant (low-high) }\\ -->
<!-- $$ -->


$$
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y})
$$
$$
Cov(x, x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(x_i - \bar{x})\\
S^2(x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2 \text{ sample variance of x}\\
S(x) = \sqrt{S^2(x)} = \sqrt{\frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2} \text{ sample standard deviation of x}\\ 
$$
$$
\rho(x, y) = \frac{Cov(x, y)}{S(x)S(y)} \text{ sample correlation between x and y}\\
-1 \leq \rho(x, y) \leq 1
$$

```{r}
## Compute the correlation between weight and height
cor(earnings$height, earnings$weight, use = "complete.obs")
cor(earnings$height_cm, earnings$weight_kg, use = "complete.obs")
```
$$
\text{Unit of measurement of the covariance?}\\
(x_i[kg] - \bar{x}[kg]) \implies (x_i - \bar{x})[kg]\\
(y_i[cm] - \bar{y}[cm]) \implies (y_i - \bar{y})[cm]\\
(x_i - \bar{x})[kg](y_i - \bar{y})[cm] => (x_i - \bar{x})(y_i - \bar{y})[cm \times kg]\\
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y})[cm \times kg]\\
S^2(x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2[kg^2] \text{ sample variance of x (weight) measured in } kg^2\\
S(x)[kg] = \sqrt{S^2(x)[kg^2]}\\
S^2(y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (y_i - \bar{y})^2[cm^2] \text{ sample variance of y (height) measured in } cm^2\\
S(y)[cm] = \sqrt{S^2(x)[cm^2]}\\
\rho(x, y) = \frac{Cov(x, y)[kg\times cm]}{S(x)[kg]S(y)[cm]}[unitless]
$$

$$
5[m] \times 4[m] = 20 [m^2]\\
2[m] \times 5[m] = 10 [m^2] \\
\frac{20 [m^2]}{10 [m^2]}[unitless] \text{ because the measurement units cancel}
$$

$$
Cov(x, x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(x_i - \bar{x}) =\\
Cov(x, x) = Var(x) = S^2(x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2\\
S(x) = \sqrt{S^2(x)} = \sqrt{\frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2}: \text{емп. стандартно отклонение, sample standard deviation}\\
S(y) = \sqrt{S^2(y)} = \sqrt{\frac{1}{n - 1}\sum_{i = 1}^{n} (y_i - \bar{y})^2}: \text{емп. стандартно отклонение, sample standard deviation}
$$
$$
\rho(x, y) = \frac{Cov(x, y)}{S(x)S(y)}: \text{ корелация между x и y}\\
-1 \leq \rho \leq 1
$$
```{r}
## Covariance (complete.obs: използвай наблюдения без липсващи стойности)
cov(earnings$weight_kg, earnings$height_cm, use = "complete.obs")

##
cor(earnings$weight_kg, earnings$height_cm, use = "complete.obs")
```

Scatter plot with no association between the variables

```{r}
## r: random, unif: uniform
# ?runif
runif(100, min = -1, max = 1)
```

```{r}
## 
df_noassoc <- data.frame(
  ## Select 100 values at random from [-1,k 1]
  ## random numbers from the uniform distribution
  x = runif(100, min = -1, max = 1),
  y = runif(100, min = -1, max = 1)
)

ggplot(data = df_noassoc, aes(x = x, y = y)) + 
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)
```
Expected correlation for the points in this plot?

```{r}
cor(df_noassoc$x, df_noassoc$y)
```

## Zero correlation

Let's say you are given a correlation coefficient of 0 for weight and height.The correlation
measures _linear_ association.


```{r}
x <- seq(0, 3 * pi, length.out = 50)
y <- sin(x)
df_nonlinear <- data.frame(x = x, y = y)


ggplot(data = df_nonlinear, aes(x = x, y = y)) +
  geom_point()

cor(x, y)
## 
```


$$
\text{e: exponent}\\
3.517357e-16 = \frac{3.517357}{10 ^ {16}}\\
10^6 = 1,000,000\\
10^9 = 1,000,000,000
$$




































## Further reading

[R for data science](https://r4ds.had.co.nz/index.html)