---
title: "Week 5"
author: "Boyko Amarov"
date: "4/19/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
kids <- read_csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2020-solutions/master/data/childiq.csv") %>% select(kid_score, mom_hs)
```

## Linear regression model with a single categorical predictor


Variables description:


- `kid_score`: (numeric) Kid’s IQ score.
- `mom_hs` (binary): 1 if the mother has finished high school, 0 otherwise.


Two groups of kids: the ones whose mother had a high school degree (`mom_hs = 1`) and the rest (`mom_hs = 0`).

Question: are this two groups different with respect to IQ (as measured by `kid_score`).


```{r}
# == (logical comparison)
kids_mom_nohs <- kids %>% filter(mom_hs == 0)
kids_mom_hs <- kids %>% filter(mom_hs == 1)
```

```{r}
print("Summary of the no HS degree group")
summary(kids_mom_nohs$kid_score)

print("Summary of the with HS degree group")
summary(kids_mom_hs$kid_score)
```

```{r}
ggplot(data = kids, aes(x = mom_hs, y = kid_score, color = factor(mom_hs))) + 
  geom_point(position = "jitter")
```


```{r}
ggplot(data = kids, aes(x = kid_score, y = factor(mom_hs))) + 
  geom_boxplot()
```


Assume that the kids were selected at random from the population of children in the USA (at the time of the survey).




































## Summarise the data

```{r}
summary(kids$kid_score)
```
```{r}
table(kids$mom_hs)
```

```{r}
## select only the observations (kids) whose mother had a high school degree
## == (logical comparison)
kids_mom_nohs <- kids %>% filter(mom_hs == 0)
kids_mom_hs <- kids %>% filter(mom_hs == 1)
```

```{r}
print("Kids with mothers without high school degree")
summary(kids_mom_nohs$kid_score)
print("Kids with mothers with high school degree")
summary(kids_mom_hs$kid_score)
```
```{r}
ggplot(data = kids, aes(x = mom_hs, y = kid_score, color = factor(mom_hs))) + 
  geom_point(position = "jitter")
```

```{r}
ggplot(data = kids, aes(x = kid_score, y = factor(mom_hs))) + 
  geom_boxplot()
```

Assume that the kids were selected at random from the population of all kids in the USA.


There are a lot of samples that we could have observed, but we have selected only _one_ sample!

Assumption about all the possible samples. Before that we formalise our comparison between the two groups
in terms of a linear regression model.

$$
i = 1,\ldots,n = 434 \text{ kids}\\
y_i: \text{ IQ score of kid } i\\
x_i: \text{ mother has high school degree (1) or not (0) for kid } i\\
$$
$$
y_i = \alpha + \beta x_i + e_i, \quad E(e_i) = 0\\
E(y_i) = \alpha + \beta x_i\\
\begin{align}
E(y_i) = 
\begin{cases} 
\mu_1 = \alpha + \beta & \text{if } x_i = 1 \\
\mu_0 =\alpha & \text{if } x_i = 0
\end{cases}   
\end{align}
$$


In the linear model $\alpha$ is the the expected IQ score of kids with mother without high school degree. The expected IQ score of kids with mother with HS degree is $\alpha + \beta$.



$$
\mu_1 = \alpha + \beta\\
\mu_0 =\alpha\\
\implies \beta = \mu_1 - \mu_0
$$
$\beta$ is the difference between expected IQ scores of kids in the first group (mother has HS degree) and expected IQ score of kids in the second group (kids with mother without HS degree).

## Coefficient estimation


```{r}
lm(kid_score ~ 1 + mom_hs, data = kids)
```
Estimated regression equation
$$
\hat{\alpha} = 77.55 \\
\hat{\beta} = 11.77 \\
\hat{y}_i = \hat{\alpha} + \hat{\beta}x_i\\
$$

$$
\hat{y}_i = \underbrace{77.55}_{\text{sample average of the x = 0 group}} + \underbrace{11.77}_{89.32 - 77.55: \text{difference between the group sample averages}}x_i\\
$$
## Normal distribution

The density of the normal distribution is given by:
$$
f(x) = \frac{1}{\sqrt{\pi \sigma^2}}e^{-\frac{1}{2}\left(\frac{x - \mu}{\sigma}\right)^2}, \quad y \in R
$$
for short we write $N(\mu, \sigma^2)$. The normal distribution with mean $\mu = 0$ and standard deviation $\sigma = 1$ is called the _standard_ normal distribution. You don't have to remember the formula, thou. You can compute its value using the `dnorm` function, for example:

```{r}
## Value of the density with mu = 0, sigma = 1: N(0, 1)
dnorm(0, mean = 0, sd = 1)

## Value of the density with mu = 5, sigma = 2: N(0, 4)
dnorm(1, mean = 5, sd = 2)
```
Visualise the density
```{r}
norm_df <- expand_grid(
  x = seq(-10, 10, length.out = 300),
  mu = c(0, 3),
  sd = c(1, 3)
) %>%
  mutate(
    y = dnorm(x, mean = mu, sd = sd)
  )
```

Visualise the densitites

```{r}
norm_df %>%
  ggplot(aes(x = x, y = y, color = factor(mu), lty = factor(sd))) +
  geom_line()
```
The area under the density over an interval equals the probability of an event occurring in the interval.

```{r}
d <- data.frame(
  x = seq(-5, 5, length.out = 200)
) %>%
  mutate(
    ## Dnorm computes the density 
    y = dnorm(x, mean = 0, sd = 1)
  )

std_norm_plot <- d %>% 
  ggplot(aes(x = x)) +
  # note this next line is the only difference in code from the last plot
  geom_ribbon(
    data = d %>% filter(x > -1, x < 1),
    aes(ymin = 0, ymax = y), fill = "gray"
  ) +
  geom_line(aes(y = y)) +
  annotate(geom = "text", x = 0, y = 0.2, label = expression("p" %~~% "??")) +
  labs(
    x = "x",
    y = "Normal density"
  )

plot(std_norm_plot)
```
Draw 100 values from the N(0, 1) distribution.
```{r}
norm_sim <- data.frame(
  x = rnorm(100, mean = 0, sd = 1)
) %>%
  mutate(
    x_in_interval = x < 1 & x > -1
  )

## Code for illustration only
std_norm_plot + 
  geom_point(
    data = norm_sim, 
    aes(x = x, y = runif(100, -0.02, 0.02), color = x_in_interval),
    size = 1/2,
    alpha = 1/2
  ) +
  geom_density(data = norm_sim, color = "steelblue2")
```
You can compute the probability for an interval using `pnorm`.

```{r}
pnorm(0, mean = 0, sd = 1)
```

$$
P(x < 0|\mu = 0, \sigma = 1) = \text{pnorm}(0, \text{mean} = 0, \text{sd} = 1) = 0.5
$$
Count how many of the simulated values are less than zero:

```{r}
## x < 0 returns a logical vector with TRUE/FALSE values
## The sum of all these values equals the number of TRUE
sum(norm_sim$x < 0)
```
```{r}
## Probability that x > 0
1 - pnorm(0, mean = 0, sd = 1)
sum(norm_sim$x > 0)
```
```{r}
## Probability that x > 0 and x < 1
pnorm(1, mean = 0, sd = 1) - pnorm(0, mean = 0, sd = 1)
sum(norm_sim$x < 1 & norm_sim$x > 0)
```

```{r}
d <- data.frame(
  x = seq(-5, 5, length.out = 200)
) %>%
  mutate(
    ## Dnorm computes the density 
    y = dnorm(x, mean = 0, sd = 1)
  )

## Code for illustration only
d %>% 
  ggplot(aes(x = x)) +
  # note this next line is the only difference in code from the last plot
  geom_ribbon(data = d %>% filter(x > -1, x < 1),
              aes(ymin = 0, ymax = y), fill = "gray") +
  geom_line(aes(y = y)) +
  annotate(
    geom = "text", 
    x = 0, 
    y = 0.2, 
    label = substitute(
      "Area" %~~% p,
      list(
        p = round(pnorm(1) - pnorm(-1), 2)
      )
    )
  ) +
  labs(
    x = "x",
    y = "Normal density"
  )
```

```{r}
d %>% 
  ggplot(aes(x = x)) +
  # note this next line is the only difference in code from the last plot
  geom_ribbon(data = d %>% filter(x > 1.96),
              aes(ymin = 0, ymax = y), fill = "gray") +
  geom_line(aes(y = y)) +
  annotate(
    geom = "text", 
    x = 3,
    y = 0.05, 
    label = substitute(
      "Area" %~~% p,
      list(
        p = round(1 - pnorm(1.964), 3)
        )
      )
    ) +
  labs(x = "x",
       y = "Normal density")
```


```{r}

d <- data.frame(
  x = seq(-5, 5, length.out = 200)
) %>%
  mutate(
    ## Dnorm computes the density 
    y = dnorm(x, mean = 0, sd = 1)
  )

d %>% 
  ggplot(aes(x = x)) +
  # note this next line is the only difference in code from the last plot
  geom_ribbon(data = d %>% filter(x > -1, x < 1),
              aes(ymin = 0, ymax = y), fill = "gray") +
  geom_line(aes(y = y)) +
  annotate(geom = "text", x = 0, y = 0.2, label = expression("p" %~~% "??")) +
  labs(x = "x",
       y = "Normal density")
```

## Simulation

Let us assume that the $e_i$ term follows a normal distribution with mean zero and standard deviation $\sigma$.

```{r}
## Do the same calculation of the group averages that
## we accomplished by splitting the data in two using `filter`.
kids %>% 
  group_by(mom_hs) %>%
  summarise(
    n = n(),
    mean = mean(kid_score)
  )

## Estimate the standard deviation of the IQ scores (for all groups)
## so that we can base our simulation on a plausible value
sd(kids$kid_score)

## Number of simulated samples

## Construct a data set with mom_hs repeated R = 2000 times
## 
sim_df <- expand_grid(
    R = 1:2000,
    mom_hs = kids$mom_hs
  ) %>%
  mutate(
    ## Compute a simulated IQ score for each kid according to our estimated regression equation
    ## rnorm adds a value selected at random from a normal distribution with mean = 0 and standard
    ## deviation (sigma) = 20.41 (that we estimated from the sample)
    kid_score = 77.54839 + (89.31965 - 77.54839) * mom_hs + rnorm(2000 * 434, mean = 0, sd = 20.41069)
  )
```
```{r}
sim_df %>%
  filter(R < 50) %>%
  ggplot(
    aes(
      x = kid_score, 
      group = interaction(R, mom_hs),
      colour = factor(mom_hs)
      )
    ) +
  geom_density(
    size = 1/5
  )
```

Estimate the regression coefficients from each sample

```{r}
# tidy(lm(kid_score ~ mom_hs, data = kids))

sim_coeff <- sim_df %>%
  group_by(R) %>%
  ## The tidy function reformats the output of lm so that it can fit in a data frame
  do(tidy(lm(kid_score ~ mom_hs, data = .)))
```


Plot the simulated distribution of the beta coefficient (difference between the group sample averages).
```{r}
sim_coeff %>%
  filter(term == "mom_hs") %>%
  ggplot(aes(x = estimate)) + 
  geom_density()
```

```{r}
## Compute the mean of each column by coefficient
sim_coeff %>%
  group_by(term) %>%
  summarise_all(mean)
```

Estimated regression equations for each simulated sample

```{r}
sim_coeff_wide <- sim_coeff %>%
  mutate(
    ## Rename the terms to get easier variable names
    term = ifelse(term == "(Intercept)", "alpha", "beta")
  ) %>%
  select(R, term, estimate) %>%
  ## Pivot the data set so that we have the estimated coefficients as columns
  pivot_wider(names_from = "term", values_from = "estimate")
```

Draw the straight lines estimated for each simulated sample.

```{r}
sim_coeff_wide %>%
  filter(R < 50) %>%
  ggplot() + 
    geom_abline(aes(intercept = alpha, slope = beta, group = R), alpha = 0.1, size = 1/5) + 
    ## Set the limits of x-axis
    xlim(c(0, 1)) +
    ## Set the limits of y-axis
    ylim(c(0, 140))
```





