---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---

## Data import

```{r, setup, message=FALSE}
## Load packages
library(tidyverse)
## Read the data
earnings <- read.csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```


```{r}
glimpse(earnings)
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years

Create a new column holding the height in cm. We use the `mutate` function for this task.

```{r}
##
## earnings <- mutate(earnings, height_cm = 2.54 * height)
## Equivalently using pipes:
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

```{r}
mean(earnings$height_cm)
```
The average height of the persons was 169 cm.

```{r}
summary(earnings$height_cm)
```
```{r}
mean(earnings$height_cm < 162.6)
```
The tallest person was 208 cm tall. The shortest person was 144.8 cm. Median equals 167.6 cm meaning that: _about_ half of the persons
in the survey were shorter than 167.6 cm (median/2nd quartile/0.5 quantile/50% percentile). 1st quartile equals 162.6: _roughly_ (because of rounding in height value) one quarter (25%) were shorter than 162.6 cm. 3th quartile equals 175.9 cm: _about_ 75% (three quarters) of the persons were shorter than 175.9 cm.

## Box-Whiskers plot

```{r}
ggplot(data = earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```
```{r}
ggplot(data = earnings, aes(x = height_cm, y = factor(male))) + geom_boxplot()
```

```{r}
## Extreme case of no variation in the data (all values equal)
summary(c(2, 2, 2, 2, 2, 2, 2))
ggplot(data = data.frame(x = c(2, 2, 2, 2, 2, 2, 2)), aes(x = x)) + geom_boxplot()
```

1. Location: (measured by the median)
2. Spread: (distance between min/max and between the 1st and the 3the quartiles)
3. Outliers: observations that are different from the rest of the observations (persons)

For _categorical_ variables we use frequency tables as a summary.

```{r}
table(earnings$ethnicity)
```

```{r}
ggplot(data = earnings, aes(x = ethnicity)) + geom_bar()
```

There are 180 black persons in the dataset.

## Measurement scales

```{r}
earnings <- earnings %>%
  mutate(
    sex = ifelse(male == 1, "Male", "Female")
  )
```

Categorical scales:

1. **Nominal**: Values of the variable have no numeric meaning. Values are only labels indicating group membership.
2. **Ordinal**: Values of the variable are just labels (as is the case with the nominal scale), but the labels are ordered.
 school grades
 "2" "3" "4" "5" "6"

Numeric scales
<!-- 3. **Interval**: -->
4. **Ratio scale**: Numeric scale, ratios between values make sense.


## Association

Scatter plot

```{r}
## Compute the mean by simply omitting the missing values
mean(earnings$weight_kg, na.rm = TRUE)

mean(earnings$height_cm)

## NA: missing/undefined value
```
```{r}
NA + 2
```


```{r}
ggplot(
  data = earnings, 
  ## aes: aesthetics
  mapping = aes(x = weight_kg, y = height_cm)
  ) +
  geom_point(position = "jitter", alpha = 0.3) +
  geom_hline(yintercept = 169) + 
  geom_vline(xintercept = 70)
```

Scatter plot with no association between the variables

```{r}
## r: random, unif: uniform
# ?runif
runif(100, min = -1, max = 1)
```

```{r}
## 
df_noassoc <- data.frame(
  ## Select 100 values at random from [-1,k 1]
  ## random numbers from the uniform distribution
  x = runif(100, min = -1, max = 1),
  y = runif(100, min = -1, max = 1)
)

ggplot(data = df_noassoc, aes(x = x, y = y)) + 
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)
```

## Further reading

[R for data science](https://r4ds.had.co.nz/index.html)