---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---

## Data import

```{r, setup, message=FALSE}
## Load packages
library(tidyverse)
## Read the data
earnings <- read.csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```


```{r}
glimpse(earnings)
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years

## Basic summaries

```{r}
mean(earnings$height)
```
The persons were 66.5 inches tall (on average).

Before we continue, let us create a new column `height_cm` in the dataset `earnings`.

```{r}
# earnings <- mutate(earnings, height_cm = 2.54 * height)

## Equivalently using pipe syntax
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height
  )
```

```{r}
mean(earnings$height_cm)
```
```{r}
summary(earnings$height_cm)
```
Tallest person was 208.3 cm tall (Max).  The shortest person was 144.8 cm tall (Min).
About half of the persons were shorter than 167.6 cm (Median/2nd quartile/0.5 quantile). About half are taller than 167.6 cm (Median).

1st quartile 162.6 cm: About one quarter (25%) of the persons were shorter than 162.6 cm (about 75% were taller).


### Box-Whiskers Plot

```{r}
ggplot(data = earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```

1. Location (median)
2. Spread (diversity/heterogeneity/variability)

```{r}
## Example with identical values: min = max
summary(c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1))
```

For _categorical_ variables we use frequncy tables as a summary.

```{r}
table(earnings$ethnicity)
```
```{r}
ggplot(data = earnings, aes(x = ethnicity)) + geom_bar()
```

There are 180 black persons in the dataset.

## Measurement scales

```{r}
earnings <- earnings %>%
  mutate(
    sex = ifelse(male == 1, "Male", "Female")
  )
```

Categorical scales:

1. **Nominal**: Values of the variable have no numeric meaning. Values are only labels indicating group membership.
2. **Ordinal**: Values of the variable are just labels (as is the case with the nominal scale), but the labels are ordered.
 school grades
 "2" "3" "4" "5" "6"

Numeric scales
<!-- 3. **Interval**: -->
4. **Ratio scale**: Numeric scale, ratios between values make sense.

