---
title: "Week4"
author: "Boyko Amarov"
date: "3/30/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
invoices <- read.delim('https://raw.githubusercontent.com/feb-uni-sofia/econometrics2020-solutions/master/data/invoices.txt')
```


```{r}
glimpse(invoices)
```

Variables description:

- `Day` (numeric): day
- `Invoices` (numeric): number of invoices
- `Time` (numeric): Time needed to process the invoices (hours)

Task is to predict the time needed to process 50, 120, 200, 250 invoices.

```{r}
summary(invoices$Time)
```


```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_vline(xintercept = c(80, 250), lty = 2, alpha = 0.3) +
  geom_vline(xintercept = 120, lty = 2) +
  geom_hline(yintercept = 2.110) +
  geom_abline(intercept = 0.1, slope = 0.015, colour = "steelblue2")
```

Estimate the expected processing time for 80, 120 and 250 invoices (3 estimates).

Strategy 1: Use the average processing time for the predictions

$$
i = 1,\ldots,n = 30\\
x_i: \text{ number of invoices on day } i \\
y_i: \text{ processing time on day } i \\
$$
Assume a simple (linear) functional relationship

$$
y = 0.1 + 0.015x
$$



$$
y[\text{hours}] = 0.1[\text{hours}] + \underbrace{0.015\left[\frac{\text{hours}}{\text{invoice}}\right] x[\text{numb. invoices}]}_{\text{hours}}\\
x = 0 \implies y [\text{hours}] = 0.1 [\text{hours}]
$$


$$
50 [m^2] \text{ room}\\
\text{rent } 2[\frac{BGN}{m^2}]\\
2 [\frac{BGN}{m^2}] \times 50 [m^2] = 100 [BGN]
$$

$$
y[\text{costs}] = 0.1[\text{fixed costs}] + 0.015[\text{marginal cost}]x[\text{production}]
$$

$$
y_{x = 120} = 0.1 + 0.015 \times 120 = 1.9 [\text{hours}]
$$

$$
y = \underbrace{0.1}_{\text{intercept}} + \underbrace{0.015}_{\text{slope}}x
$$

$$
y_i  = \underbrace{0.1 + 0.015x_i}_{\text{line, systematic part}} + \underbrace{e_i}_{\text{deviation from the line, stochastic (random) part}}
$$



```{r}
x <- 190 ## Number of invoices
e <- runif(1, min = -1, max = 1) ## Uniform distribution on [-1, 1]
e
y <- 0.1 + 0.015 * x + e
y
```

```{r}
rand_numbers <- runif(10, min = -1, max = 1)
# rand_numbers
paste("Average result from a game:", mean(rand_numbers))
rand_numbers
```

Assume that the deviation part is zero on average.

$$
E(e_i) = 0 \text{ assumption} \implies \\
E(y_i) = E(0.1 + 0.015x_i + e_i) =\\ 
E(y_i) = E(0.1) + E(0.015x_i) + E(e_i) = \\
E(y_i) = 0.1 + 0.015x_i
$$

## Coefficient estimation

$$
y = 0.1 + 0.015x + e_i
$$
Predicted y (expected value of y) for every observation

$$
\hat{y}_i = 0.1 + 0.015x_i
$$


```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_abline(intercept = 0.6, slope = 0.015, colour = "steelblue2") +
  geom_abline(intercept = 0.3, slope = 0.02, colour = "steelblue2", alpha = 0.5) +
  geom_abline(intercept = 0.6, slope = 0.010, colour = "steelblue2", alpha = 0.5) +
  geom_abline(intercept = 2, slope = 0.015, colour = "firebrick")
```


$$
y_i - \hat{y_i} \text{ residual for day } i
$$
```{r}
## Illustration only
invoices <- invoices %>%
  mutate(
    y_predicted_1 = 0.1 + 0.015 * Invoices
  )

ggplot(data = invoices, aes(x = Invoices, y = Time)) +
  geom_point(size = 2) +
  geom_segment(
    aes(xend = Invoices, yend = y_predicted_1),
    size = 1 / 4
  ) +
  geom_abline(intercept = 0.1, slope = 0.015, color = "steelblue")
```
How to aggregate each individual deviation (residual)? One possible approach is to simply sum the deviations

$$
\sum_{i = 1}^{n} (y_i - \hat{y}_i)^2
$$


$$
y_i = \alpha + \beta x_i + e_i\\
E(y_i) = \alpha + \beta x_i
$$


Find guesses for $\alpha$ and $\beta$ based on the data, so that the distance between
the predictions and the observations is as small as possible. Let's call our
"guesses" for $\alpha$ $\hat{\alpha}$ and for $\beta$ $\hat{\beta}$

$$
\hat{y}_i = \hat{E}(y_i) = \hat{\alpha} + \hat{\beta}x_i \text{ guessed expected value}\\
(\hat{\alpha}^{OLS}, \hat{\beta}^{OLS}) = \min_{\hat{\alpha}, \hat{\beta}} \sum_{i = 1}^{n}(y_i - \hat{y}_i)^2
$$

The result of this minimisation is what we call the OLS (ordinary least squares) estimator for $\alpha$ and for $\beta$.

## Application in R

$$
\underbrace{y_i}_{Time} = \alpha + \beta \underbrace{x_i}_{Invoices} + e_i
$$

```{r}
## lm: linear model
lm(Time ~ 1 + Invoices, data = invoices)
```

Estimated equation using OLS.

$$
\hat{\alpha} = 0.64\\
\hat{\beta} = 0.011\\
\hat{y}_i =  0.64 + 0.011 x_i
$$
We estimate the expected processing time at $x_i$ invoice to be the sum the estimated fixed costs (0.64 hours) plus the estimated marginal costs (0.011 hours/invoice) times the number of invoices ($x_i$).

The expected processing time increases by 0.011 *hours* for each additional invoice.

The expected processing time at zero invoices is 0.64 _hours_.

## Plot the OLS estimated line

```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_abline(intercept = 0.1, slope = 0.015, colour = "steelblue2") + 
  geom_abline(intercept = 0.64, slope = 0.011, colour = "firebrick")
```

```{r}
ggplot(data = invoices, aes(x = Invoices, y = Time)) + 
  geom_point() +
  geom_smooth(method = "lm")
```
