---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---


---
title: 'Week2: basic data summaries'
author: "Boyko Amarov"
date: "3/8/2021"
output: html_document
---

## Data import

```{r, setup, message=FALSE}
## Load packages
library(tidyverse)
## Read the data
earnings <- read.csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```


```{r}
glimpse(earnings)
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years

Create a new column holding the height in cm. We use the `mutate` function for this task.

```{r}
##
## earnings <- mutate(earnings, height_cm = 2.54 * height)
## Equivalently using pipes:
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

```{r}
mean(earnings$height_cm)
```
The average height of the persons was 169 cm.

```{r}
summary(earnings$height_cm)
```
```{r}
mean(earnings$height_cm < 162.6)
```
The tallest person was 208 cm tall. The shortest person was 144.8 cm. Median equals 167.6 cm meaning that: _about_ half of the persons
in the survey were shorter than 167.6 cm (median/2nd quartile/0.5 quantile/50% percentile). 1st quartile equals 162.6: _roughly_ (because of rounding in height value) one quarter (25%) were shorter than 162.6 cm. 3th quartile equals 175.9 cm: _about_ 75% (three quarters) of the persons were shorter than 175.9 cm.

## Box-Whiskers plot

```{r}
ggplot(data = earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```
```{r}
ggplot(data = earnings, aes(x = height_cm, y = factor(male))) + geom_boxplot()
```

```{r}
## Extreme case of no variation in the data (all values equal)
summary(c(2, 2, 2, 2, 2, 2, 2))
ggplot(data = data.frame(x = c(2, 2, 2, 2, 2, 2, 2)), aes(x = x)) + geom_boxplot()
```

1. Location: (measured by the median)
2. Spread: (distance between min/max and between the 1st and the 3the quartiles)
3. Outliers: observations that are different from the rest of the observations (persons)

For _categorical_ variables we use frequency tables as a summary.

```{r}
table(earnings$ethnicity)
```

```{r}
ggplot(data = earnings, aes(x = ethnicity)) + geom_bar()
```

There are 180 black persons in the dataset.

## Measurement scales

```{r}
earnings <- earnings %>%
  mutate(
    sex = ifelse(male == 1, "Male", "Female")
  )
```

Categorical scales:

1. **Nominal**: Values of the variable have no numeric meaning. Values are only labels indicating group membership.
2. **Ordinal**: Values of the variable are just labels (as is the case with the nominal scale), but the labels are ordered.
 school grades
 "2" "3" "4" "5" "6"

Numeric scales
<!-- 3. **Interval**: -->
4. **Ratio scale**: Numeric scale, ratios between values make sense.


## Association

Scatter plot

```{r}
NA + 2
```


```{r}
## Compute the mean by simply omitting the missing values
mean(earnings$weight_kg, na.rm = TRUE)
mean(earnings$height_cm)
## NA: missing/undefined value
```


```{r}
ggplot(
  data = earnings, 
  ## aes: aesthetics
  mapping = aes(
    x = weight_kg, 
    y = height_cm
    )
  ) +
  ## Add points
  geom_point(
    ## Add some random noise to all points to avoid overplotting
    position = "jitter",
    ## Transparency
    alpha = 0.25
  ) +
  ## Add horizontal line at the average height (cm)
  geom_hline(yintercept = 169, color = "steelblue2") + 
  ## Add vertical line at the average weight (kg)
  geom_vline(xintercept = 70, color = "firebrick4")  +
  ## Add labels
  labs(
    x = "Weight (kg)",
    y = "Height (cm)",
    title = "Scatterplot weight and height"
  )
```

From the points in the scatter-plot we see that heavier persons also tended to be taller persons.


$$
i = 1,\ldots,n = 1816 \text{ persons}\\
x_i: \text{ weight of person } i\\
y_i: \text{ height of person } i\\
\text{Avearge values}:\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^n x_i\\
\bar{y} = \frac{1}{n}\sum_{i = 1}^n y_i\\
$$
$$
(x_i - \bar{x}) \text{ the deviation of x_i (weight of person i) from the average weight}\\
(y_i - \bar{y}) \text{ the deviation of y_i (height of person i) from the average height}\\
(x_i - \bar{x})(y_i - \bar{y}) \text{ cross product of deviation from the means for perons i}
$$


$$
\text{Cross product of deviations from the means}\\
(x_i - \bar{x})(y_i - \bar{y})\\
(x_i - \bar{x})(y_i - \bar{y}) \geq 0 \text{ for persons in the upper right part of the plot: high-high}\\
(x_i - \bar{x})(y_i - \bar{y}) \geq 0 \text{ for persons in the lower left of the plot: low-low}\\
(x_i - \bar{x})(y_i - \bar{y}) \leq 0 \text{ for persons in the upper left of the plot: low-high}\\
(x_i - \bar{x})(y_i - \bar{y}) \leq 0 \text{ for persons in the upper left of the plot: high-low}
$$
$$
\sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y}) = \\
\sum_{i \in \text{low-low}}\underbrace{(x_i - \bar{x})(y_i - \bar{y})}_{+} +
\sum_{i \in \text{high-high}}\underbrace{(x_i - \bar{x})(y_i - \bar{y})}_{+} +\\ 
\sum_{i \in \text{low-high}}\underbrace{(x_i - \bar{x})(y_i - \bar{y})}_{-} + 
\sum_{i \in \text{high-low}}\underbrace{(x_i - \bar{x})(y_i - \bar{y})}_{-}
$$



## Sample Covariance

Sum of the cross-products over all observation (persons).

$$
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y}) \text{ the sign of this sum depends on the arrangement of the points}
$$
```{r}
## Compute the covariance between height_cm and weight_cm
## Omit missing values for weight: use complete obs instructs it to use only 
## the persons with available height and weight measurements (no missing)

cov(earnings$height_cm, earnings$weight_kg, use = "complete.obs")
```

```{r}
cov(earnings$height, earnings$weight, use = "complete.obs")
```


```{r}
cov(earnings$height_cm, 1000 * earnings$weight_kg, use = "complete.obs")
```

$$
\text{Room area: }
2[m] \times 5[m] = 10[m \times m = m^2]
$$
Unites of measurements of the covariance

$$
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y})\\
(x_i - \bar{x})[kg]\times (y_i - \bar{y})[cm] = [kg \times cm] \text{ Unit of measurement for the covariance in this example}
$$


$$
Cov(x, x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(x_i - \bar{x})\\
S^2(x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2 \text{ sample variance of } x\\
S(x) = \sqrt{S^2(x)} \text{ sample standard deviation of } x\\
S(y) = \sqrt{S^2(y)} \text{ sample standard deviation of } y\\
\rho(x, y) = \frac{Cov(x, y)[kg \times cm]}{S(x)[kg]S(y)[cm]} \text{ sample correlation between x and y}
$$
$$
-1 \leq \rho(x, y) \leq 1
$$

```{r}
## Compute the correlation between weight and height
cor(earnings$height_cm, earnings$weight_kg, use = "complete.obs")
cor(earnings$height, earnings$weight, use = "complete.obs")
```

```{r}
## r: random, unif: uniform (равномерно)
runif(100, min = -1, max = 1)
```


```{r}
## 
df_noassoc <- data.frame(
  ## Select 100 values at random from [-1,k 1]
  ## random numbers from the uniform distribution
  x = runif(100, min = -1, max = 1),
  y = runif(100, min = -1, max = 1)
)
ggplot(data = df_noassoc, aes(x = x, y = y)) +
  geom_point() +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0)

# Expected correlation for the points in this plot?
# cov(df_noassoc$x, df_noassoc$y)
cor(df_noassoc$x, df_noassoc$y)
```
Perfect linear dependency (positive slope)

```{r}
x <- 1:10
y <- 500 + 0.1 * x

ggplot(data = data.frame(x = x, y = y), aes(x = x, y = y)) +
  geom_point() +
  geom_abline(intercept = 1, slope = 2)

cor(x, y)
```



```{r}
x <- 1:10
y <- 1 - 2 * x

ggplot(data = data.frame(x = x, y = y), aes(x = x, y = y)) +
  geom_point() +
  geom_abline(intercept = 1, slope = -2)

cor(x, y)
```

## Zero correlation

Let's say you are given a correlation coefficient of 0 for weight and height.The correlation
measures _linear_ association.

```{r}
## pi: 3.14
x <- seq(0 * pi , 1 * pi, length.out = 50)
y <- sin(x)

df_nonlinear <- data.frame(x = x, y = y)
ggplot(data = df_nonlinear, aes(x = x, y = y)) +
  geom_point()

cor(x, y)
## 
```
$$
\text{e: exponent}\\
3.517357e-16 = \frac{3.517357}{10 ^ {16}}\\
10^6 = 1,000,000\\
10^9 = 1,000,000,000
$$

## Further reading

[R for data science](https://r4ds.had.co.nz/index.html)

























## Data import

```{r, setup, message=FALSE}
## Load packages
library(tidyverse)
## Read the data
earnings <- read.csv("https://raw.githubusercontent.com/feb-uni-sofia/econometrics2021/main/data/earnings.csv")
```


```{r}
glimpse(earnings)
```

Data description

- _height_ (numeric): height in inches (1 inch = 2.54 cm)
- _weight_ (numeric): weight in pounds (1 pound $\approx$ 0.45 kilograms)
- _male_ (numeric): 1 if a person is male, 0 otherwise
- _earn_ (numeric): annual income in USD
- _earnk_ (numeric): annual income in 1000 USD
- _ethnicity_ (character): ethnicity
- _age_ (numeric): Age of the person in years


## Height

```{r}
mean(earnings$height) * 2.54
```
Средният ръст на хората в изследването е 66.5 инча (169 см).

Добавяме колонка `height_cm` в таблицата `earnings`.

```{r}
# earnings <- mutate(
#   earnings,
#   height_cm = 2.54 * height,
# )
## Еквивалентно на горния код
earnings <- earnings %>%
  mutate(
    height_cm = 2.54 * height,
    weight_kg = 0.45 * weight
  )
```

```{r}
mean(earnings$height_cm)
```

```{r}
summary(earnings$height_cm)
```

Най-ниският човек в изследването е бил висок 144.8 см (min). Най-високият човек е бил висок
208.3 см (max). _Около_ половината от хората в извадката са с ръст по-нисък от 167.6. _Около_ една четвърт от хорта
са били по-ниски от 162.5 см (1st quartile). 

1st Qu (първи квартил)
Median (втори квартил, 0.5 квантил)
3th Qu (трети квартил)

```{r}
mean(earnings$height_cm <= 167.6)
```

```{r}
ggplot(earnings, aes(x = height_cm, y = ethnicity)) + geom_boxplot()
```

1) Location (център на данните)
2) Spread (variability/diversity)

## 

Categorical scales:

Номинална скала: стойностите са само етикети,
които показват принадлежност към някаква група (ethnicity, gender, etc..)

Ординална скала: като номиналната скала, но с ред дефиниран между етикетите

"2", "3", "4", "5", "6"

```{r}
earnings %>%
  mutate(
    sex = ifelse(male == 1, "Male", "Female")
  ) %>%
  select(male, sex)
```

```{r}
mean(earnings$male)
```
```{r}
## Frequency table
table(earnings$ethnicity)
```

Bar charts
```{r}
ggplot(earnings, aes(x = ethnicity)) + geom_bar()
```

## Association

```{r}
mean(earnings$height_cm)
## Средна стойност, като преди това премахва липсващи стойности
mean(earnings$weight_kg, na.rm = TRUE)
```


Scatter plot

```{r, fig.cap=""}
## aes: aesthetics
ggplot(data = earnings, aes(x = weight_kg, y = height_cm)) +
  geom_point(position = "jitter", alpha = 0.2) +
  geom_hline(yintercept = 169, color = "steelblue2") +
  geom_vline(xintercept = 70, color = "firebrick")
```
По-тежките хора като тенденция са и по-високи на ръст.

$$
\underbrace{(x_i - \bar{x})}_{\geq 0}\underbrace{(y_i - \bar{y})}_{\geq 0} \geq 0: (\text{горе дясно})\\
\underbrace{(x_i - \bar{x})}_{\leq 0}\underbrace{(y_i - \bar{y})}_{\leq 0} \geq 0: (\text{долу ляво})\\
\underbrace{(x_i - \bar{x})}_{\leq 0}\underbrace{(y_i - \bar{y})}_{\geq 0} \leq 0: (\text{горе ляво})\\
\underbrace{(x_i - \bar{x})}_{\geq 0}\underbrace{(y_i - \bar{y})}_{\leq 0} \leq 0: (\text{горе ляво})\\
$$
$$
\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y}): \text{ знакът на този израз зависи от подребата на точките}
$$
(Емпирична) Ко-вариация (empirical Covariance)

$$
Cov(x, y) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y})
$$

Емпирична Дисперсия (empirical, sample) variance
$$
Cov(x, x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})(x_i - \bar{x}) =\\
Cov(x, x) = Var(x) = S^2(x) = \frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2\\
S(x) = \sqrt{S^2(x)} = \sqrt{\frac{1}{n - 1}\sum_{i = 1}^{n} (x_i - \bar{x})^2}: \text{емп. стандартно отклонение, sample standard deviation}\\
S(y) = \sqrt{S^2(y)} = \sqrt{\frac{1}{n - 1}\sum_{i = 1}^{n} (y_i - \bar{y})^2}: \text{емп. стандартно отклонение, sample standard deviation}
$$
$$
\rho(x, y) = \frac{Cov(x, y)}{S(x)S(y)}: \text{ корелация между x и y}\\
-1 \leq \rho \leq 1
$$
```{r}
## Covariance (complete.obs: използвай наблюдения без липсващи стойности)
cov(earnings$weight_kg, earnings$height_cm, use = "complete.obs")

##
cor(earnings$weight_kg, earnings$height_cm, use = "complete.obs")
```



## Correlation

$$
x_i: \text{тегло в кг.}\\
y_i: \text{ръст в кг.} \\
i = 1,\ldots, n (= 1816 \text{ в настоящия пример})\\
(x_1, x_2,\ldots,x_n)\\
\bar{x} = \frac{1}{n}\sum_{i = 1}^{n} x_i\\
\bar{y} = \frac{1}{n}\sum_{i = 1}^{n} y_i\\

$$



